Reionization Planck 2018 ...

### 2020-11-04 paper/rei.tex
Updates from Overleaf

\section{Background}
\label{sec:background}

@@

\subsection{Reionization Principal Components}
\label{sec:PC}

We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PCs $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{1}{\sigma_l^2} \frac{\partial C_l^{EE}}{\partial x_e(z_i)}\frac{\partial C_l^{EE}}{\partial x_e(z_j)} = \sum_a S_a(z_i) \sigma_a^{-2} S_a(z_j),
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs. We have rank-ordered the PCs from low to high variance, so that for the range $z_{min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g. \cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ and $n_{\rm{PC}} = 7$ for $z_{\rm max} = 50$.

@@

Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.

[rephrase this paragraph:]
``For the Planck data set, most of the information in the ionization history is carried by the
first two modes and therefore relates to the amount of high vs.~low redshift optical depth. We keep all 5 PCs  for completeness in representing the observable
impact of a given ionization history and to marginalize uncertainties that they introduce."

@@

We also put constraints on the cumulative Thomson optical depth derived parameters 
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. Because the optical depth is more directly related to the CMB observables than the ionization history itself, it allows us to make meaningful comparisons between the PC results and those with a specific prior in model space such as those admitting a particular functional form.

For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
 \begin{equation}
x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
 \label{eqn:tanh}
 \end{equation}
 where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
and has an implicit prior in model space that excludes finite ionization above its transition redshift.

The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed. When showing results, however, we do impose a physicality prior derived from the condition $0<x_e<x_e^{\rm max}$ following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
%
Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality. They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.

%we employ the smoothed $x_e$ which formally has support beyond the bounds. 
% We include this small correction by integrating
% slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
   %Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
%Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.

\subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
\label{sec:KDE}

@@

The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}. Since $\mathbf{m}$ could take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.

@@

Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
\end{equation}
we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time. We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

@@

\section{Planck 2018 PC Results}
\label{sec:results}

@@

\begin{figure*}
\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries. [TODO: update legend] [add tau trajectories] [add prior boundaries]}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf

@@

\begin{figure}
\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wGaussEllipse.png}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, $m_1-m_2$ plane, comparing Planck 2018 (blue) vs 2015 results (red). The black solid line indicates the steplike models with the arrow pointing in the direction of increasing $\tau$. The dashed lines indicate the contours corresponding a Gaussian likelihood (Eq.~\ref{eq:gaussian}) using the PC covariance from the Planck 2018 chains. [add Gaussian covariance legend?] [add tau trajectories here?]
}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

\begin{table}[b]
\centering
\caption{[Placeholder, need to fill in real numbers] PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$.}
\label{tab:PC_stats}
\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
\hline

			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$} 
		\\ \hline
$m_1$ 
	& 0.002 & 0.053 & 1.000 & 0.450 & $-$0.432 & 0.273 & $-$0.073 \\ 
$m_2$ 
	& $-$0.030 &  0.101 & 0.450 & 1.000 & $-$0.262 & 0.055 & 0.072 \\ 
$m_3$ 
	& 0.019 &  0.128 & $-$0.432 & $-$0.262 &1.000 & $-$0.417 & 0.155 \\
$m_4$  
	& $-$0.012 & 0.143 &  0.273 & 0.055 & $-$0.417 & 1.000 & $-$0.428 \\ 
$m_5$ 
	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ \hline
\end{tabular}
\end{table}

\begin{figure}
\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
\caption{Same as Fig.~\ref{fig:plot_mjs_2018_vs_2015} but comparing the Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the official \texttt{simall\_EE} likelihood and the \texttt{srollv2} likelihood which has better foreground removal (shown throughout this work). [Comments on shift m1-m2 plane].
}
\label{fig:plot_m1m2_lowE_vs_srollv2}
\end{figure}

We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area. [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]

In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to an even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$ (also shown in Table~\ref{tab:PC_stats}:
\begin{equation}
 {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
 This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement. The Gaussian likelihood is even faster than the kernel density estimate, and may be used to sample well-constrained models near the peak of the PC distributions, while becoming possibly less accurate near the tails of the distributions.

@@

In Fig.~\ref{fig:plot_taugtz_lowE_vs_srollv2}, we compare the Planck 2018 results with the two different low-$\ell$ $EE likelihoods.

@@

We show however, using a separate PC chains with $\zmax = 50$, that there is no evidence for optical depth at higher redshifts than 30. In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that the contours in $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines), and that their total optical depth constraints are practically identical: 
\beq
\tau_{\rm PC, \zmax=30} = XX \pm XX,
\eeq
and
\beq
\tau_{\rm PC, \zmax=50} = XX \pm XX.
\eeq
We have an upper limit on $\tau(15, \zmax)$ of 
\beq
\tau(z, \zmax)_{\rm PC, \zmax=30} < XX \; (95\% CL)
\eeq
and 
\beq
\tau(z, \zmax)_{\rm PC, \zmax=50} < XX \; (95\% CL)
\eeq

@@

\begin{itemize}
    \item compare with Planck official results
    \item comment that Planck tau(15,30) constraints are too stringent (flex knots). Compare to ours. Point out why.
    \item{ Comments on priors (see below).}
    \item (*maybe in  another section): make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.
\end{itemize}

Recap our appendix from Planck 2015 on the issue of prior:
\begin{itemize}
    \item {describe tau prior issue raised by MB18: claimed that prior not flat in tau so bias PC; proposed to flatten by multiplying the inverse of prior point-by-point in the PC space; claim that this flattening removes a shift between tanh and PC results. }
    \item {summarize that we pointed it out the flaw in the logic in appendix of \cite{Heinrich:2018btc}.}
    \item{reproduce a few tests here: 1) tau12 posterior (no need to show again, point to appendix); 2) show prior boxes in m1-m2 plane (done in Fig.4) - repeat the length of lines of constant tau12 within the box describe the prior distribution. Make the point that though the line rises, the Planck data is more informative than priors in m1, m2 makes no impact on the posterior. However when inverting the prior, introduces a bias in tau.}
    \item{show test of explicitly constructing a prior that's locally flat in m1-m2, show posterior almost doesn't change. (only change is low-z enhancement, due to prior cut off at $z>6$. between our two priors, things don't change as much (might want to reproduce this result for this Planck 2018 results. And marius version too?}
    \item{inversion method should not be used on future dataset, since data is more informative than prior.}
\end{itemize}

 \begin{figure}
          \includegraphics[width=0.9\columnwidth]{paper/plots/plot_rotated_box_flat_tau_prior_fac_0p8.pdf}
          \caption
          {[Placeholder right now (Planck 2015 plot)] Priors on the $m_1-m_2$ parameter space: the original HMH16 prior (inset square, solid lines) and a prior that is flat in
          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) whose sides are aligned with lines of
          constant $\tau_{12}$ (light gray lines).   The HMH16 prior allows more parameter space at high vs low $\tau_{12}$ but this
          is not relevant since the data constraints (ellipses) exclude this region.  In the allowed region, the only difference is that the HMH16 prior clips the low $m_1$ edge of the allowed region due to physicality and the assumption that reionization occurs at
          $z\ge 6$. [add in text the comparison of priors, and how tau is not biased, and for KDE physical models determine the priors]}  \label{fig:prior_box}
\end{figure}

[Plot: add plot for tau(>z) contours for the 2 parameter model vs PC; direct integration first]

>>>>

\section{Background}
\label{sec:background}

@@

\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PCs $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{1}{\sigma_l^2} \frac{\partial C_l^{EE}}{\partial x_e(z_i)}\frac{\partial C_l^{EE}}{\partial x_e(z_j)} = \sum_a S_a(z_i) \sigma_a^{-2} S_a(z_j),
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs. We have rank-ordered the PCs from low to high variance, so that for the range $z_{min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g. \cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ and $n_{\rm{PC}} = 7$ for $z_{\rm max} = 50$.

@@

Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
[rephrase this paragraph:]
``For the Planck data set, most of the information in the ionization history is carried by the
first two modes and therefore relates to the amount of high vs.~low redshift optical depth. We keep all 5 PCs  for completeness in representing the observable
impact of a given ionization history and to marginalize uncertainties that they introduce."

@@

We also put constraints on the cumulative Thomson optical depth derived parameters 
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. Because the optical depth is more directly related to the CMB observables than the ionization history itself, it allows us to make meaningful comparisons between the PC results and those with a specific prior in model space such as those admitting a particular functional form.
For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
 \begin{equation}
x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
 \label{eqn:tanh}
 \end{equation}
 where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
and has an implicit prior in model space that excludes finite ionization above its transition redshift.

The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed. When showing results, however, we do impose a physicality prior derived from the condition $0<x_e<x_e^{\rm max}$ following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
%
Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality. They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.
%we employ the smoothed $x_e$ which formally has support beyond the bounds. 
% We include this small correction by integrating
% slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
   %Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
%Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.   
\subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
\label{sec:KDE}

@@

The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}. Since $\mathbf{m}$ could take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.

@@

Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
\end{equation}
we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time. We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization. 

@@

\section{Planck 2018 PC Results}
\label{sec:results}

@@

\begin{figure*}
\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries. [TODO: update legend] [add tau trajectories] [add prior boundaries]}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf

@@

\begin{figure}
\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wGaussEllipse.png}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, $m_1-m_2$ plane, comparing Planck 2018 (blue) vs 2015 results (red). The black solid line indicates the steplike models with the arrow pointing in the direction of increasing $\tau$. The dashed lines indicate the contours corresponding a Gaussian likelihood (Eq.~\ref{eq:gaussian}) using the PC covariance from the Planck 2018 chains. [add Gaussian covariance legend?] [add tau trajectories here?]
}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}
\begin{table}[b]
\centering
\caption{[Placeholder, need to fill in real numbers] PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$.}
\label{tab:PC_stats}
\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
\hline

			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$} 
		\\ \hline
$m_1$ 
	& 0.002 & 0.053 & 1.000 & 0.450 & $-$0.432 & 0.273 & $-$0.073 \\ 
$m_2$ 
	& $-$0.030 &  0.101 & 0.450 & 1.000 & $-$0.262 & 0.055 & 0.072 \\ 
$m_3$ 
	& 0.019 &  0.128 & $-$0.432 & $-$0.262 &1.000 & $-$0.417 & 0.155 \\
$m_4$  
	& $-$0.012 & 0.143 &  0.273 & 0.055 & $-$0.417 & 1.000 & $-$0.428 \\ 
$m_5$ 
	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ \hline
\end{tabular}
\end{table}
\begin{figure}
\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
\caption{Same as Fig.~\ref{fig:plot_mjs_2018_vs_2015} but comparing the Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the official \texttt{simall\_EE} likelihood and the \texttt{srollv2} likelihood which has better foreground removal (shown throughout this work). [Comments on shift m1-m2 plane].
}
\label{fig:plot_m1m2_lowE_vs_srollv2}
\end{figure}
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area. [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.] 
In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to an even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$ (also shown in Table~\ref{tab:PC_stats}:
\begin{equation}
 {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
 This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement. The Gaussian likelihood is even faster than the kernel density estimate, and may be used to sample well-constrained models near the peak of the PC distributions, while becoming possibly less accurate near the tails of the distributions.

@@

In Fig.~\ref{fig:plot_taugtz_lowE_vs_srollv2}, we compare the Planck 2018 results with the two different low-$\ell$ $EE$ likelihoods. [Do we need this?]

@@

We show however, using a separate PC chains with $\zmax = 50$, that there is no evidence for optical depth at higher redshifts than 30. In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that the contours in $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines), and that their total optical depth constraints are practically identical: 
\beq
\tau_{\rm PC,\, \zmax=30} = XX \pm XX,
\eeq
and
\beq
\tau_{\rm PC,\, \zmax=50} = XX \pm XX.
\eeq
We have an upper limit on $\tau(15, \zmax)$ of 
\beq
\tau(15, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})
\eeq
and 
\beq
\tau(15, \zmax)_{\rm PC,\, \zmax=50} < XX \; (95\% \mathrm{CL})
\eeq

@@

The Planck collaboration has released constraints on the total optical depth $\tau$ with a steplike model parameterized by the tanh function, which has evolved from $\tau = 0.067 \pm 0.022$ (Planck 2015) to $\tau = 0.055 \pm 0.009$ (Planck 2016), then to $\tau = 0.0506 \pm 0.0086$ (Planck 2018) [cite papers]. The major improvement comes from clean-up of systematics in the High Frequency Instrument (HFI) data.

In Planck 2018 VI, using only lowE data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for a steplike model, $\tau = 0.0504^{+0.0050}
_{-0.0079}$ for the FlexKnot method and $\tau = 0.0487^{+0.0038}_{−0.0081}$ for the PC method with a different prior than that used in this paper.

They also found the upper limit on high-redshift optical depth to be: $\tau(15, 30) < 0.006$ for the FlexKnot method with a flat $\tau(15, 30)$ prior or $<0.007$ with a flat knot prior. The PC result for the upper limit was not computed.

Here we show that we find a less stringent upper limit using the PC method $\tau(15, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})$. [explain why FlexKnot could give more stringent results.]

\subsection{A note on priors}

In Ref.~\cite{Millea:2018bko} (hereafter MB18)  a concern was raised that the physicality prior we take for the PC analysis in Ref.~\cite{Heinrich:2016ojb} implies a non-flat prior in $\tau$ and introduces bias to the PC result. The authors also proposed a method to flatten the $\tau$ prior by multiplying its inverse at each point in the multidimensional PC space. They claim that adopting the physicality prior used here and in Ref.~\cite{Heinrich:2016ojb} contributed to the a large fraction of the shift in the $\tau$ posterior between the steplike and PC method in the Planck 2015 data.

In the Planck 2018 paper on cosmological parameters, the reduction of optical depth from the PC results seen in the Planck 2018 data compared to the Planck 2015 data was attributed to the reduction of HFI systematics, with subdominant contribution from priors. We therefore would like to clarify the impacts of the priors we choose and illustrate explicitly that there is no bias with this choice, whereas the flattening method proposed in MB18 is actually at risk of creating a non-flat prior and biasing results.

In the following, we will use $\tau \approx \tau_{12}$ in order to illustrate the impacts of the priors more easily. We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
\beq
\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
\eeq
where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model. Because most of the contributions to $\tau$ comes from the first two components as shown in Fig. 7 of Ref.~\cite{Heinrich:2018btc} in the appendix, we can approximate it as
\beq
\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
\eeq
We also show it here in Fig.~\ref{fig:tau12} where the posterior of $\tau_{12}$ and $\tau$ (with five PCs) are compared for the Planck 2018 data.

In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray). The solid blue box corresponds to the physicality prior, whereas the dashed blue box is constructed to be aligned with the directions of the constant $\tau_{12}$ lines, so that when projected onto the $\tau_{12}$ directions, gives a true flat $\tau_{12}$ prior. The physicality prior upon projection would correspond to a linearly rising prior in $\tau_{12}$ over the region where the posterior has support. One would be eager to conclude that this produces a biased constraint on $\tau_{12}$. However, because the Planck data is more constraining than the prior in $m_1$ and $m_2$, (the ellipse occupying a much smaller area in one part of the prior box), the value of the prior at high $\tau_{12}$ becomes irrelevant for the posterior.

Using the procedure proposed by Ref.~\cite{Millea:2018bko}, inverting the prior value at each point in the PC space, would actually up-weight the high $\tau_{12}$ regions that are not supported by data, while down-weighting the regions that are. This produces an actual bias in the $\tau_{12}$ posterior toward lower optical depth.
